---
title: "Make Grm for asreml"
author: "jim"
date: "April 14, 2018"
output: html_document
---
R script to read in the relationship matrix for training data and current cycle individuals for ReFUS population.
Check rank, force to positive definite if required.
Fix line names.
Add in extra rows and columns for 'fixed' check varieties and other lines in phenotype data set but not in GRM.
Export the sorted line names that match matrix order as ped file and sorted augmented GRM file.
These can then be read by ASReml to fit the GBLUP model.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(synbreed)
library(Matrix)
```

Set working directory get libraries and import kinship matrix from tassel

```{r}
setwd("Q:/My Drive/GS_ReFUS_files/Predictions_2018_C4")
K = read.table("REFUSC3_4_join_imp_K.txt", stringsAsFactors = F, skip = 3)
line_names = K[,1]
K = as.matrix(K[,-1])
```

Fix the 17FL line names, which are exported from GBS pipeline with index information attached
```{r}
line_names = sub('^i[0-9]{2}_17FL', '17FL', line_names)
line_names = sub('_i[0-9]{2}$', '', line_names)
```


Check rank of matrix via synbreed:
```{r}
G = K
class(G) = "relationshipMatrix"
dimnames(G) = list(line_names, line_names)
summary(G)
```
Cool, it is full rank!

Check also with Matrix package
```{r}
rankMatrix(K, tol = 10E-8,
           method = "tolNorm2")
```

looks like it's close to singular for one column, matrix is rank 836 with this tolerance  
actual tolerance of ASReml is 10E-10*(TotalSS_uncorrected)  
so this is probably OK for most traits.  
  
Correcting singularity at the relationship matrix - skip for now
```{r}
#RealizedPD = nearPD(K, keepDiag = T)
#G = matrix(RealizedPD[[1]]@x, nrow = RealizedPD[[1]]@Dim[1])
#G = G + diag(0.01, nrow(G))
#attr(G, "dimnames") = RealizedPD[[1]]@Dimnames
```

Create a fake ped file, need to add the two check lines in the experiment and also the lines in the pheno file that we did not get genotype data on
```{r}
ped = data.frame(line = line_names, mom = 0, dad = 0, stringsAsFactors = F)
phenos = read.csv("Q:/My Drive/GS_ReFUS_files/Predictions_2017/All_Exp16_pheno.csv", stringsAsFactors = F)
pheno_lines = unique(phenos$source_id)
extras = sort(setdiff(pheno_lines, line_names))
extradf = data.frame(line = extras, mom = 0, dad = 0)
ped2 = rbind(ped, extradf)
str(ped2)
```


Now we need to augment the relationship matrix to include these extra lines.  
We assume that the un-genotyped lines have F = 0 and covariances = 0 relative to other lines. So, diagonal elements = 1, off-diags = 0

For the checks, we give them diagonal values of some large number so they don't influence the variance component estimate. Say diagonal = 99999, off-diag = 0
Checks present on the training set: FR1064, FR615, GEMS0002, M141A and NC358

```{r}
N = dim(G)[1] #number of lines in original K
Nx = length(extras) #number of extra lines
Nchecks = 5 #number of checks
Nt = N + Nx #(total number of lines in augmented K)
Kaug = matrix(data = 0, nrow = Nt, ncol = Nt)
Kaug[1:N, 1:N] = G #fill in first N rows and columns with real data
diag(Kaug[(N+1):Nt, (N+1):Nt]) = 1
diag(Kaug[(Nt-Nchecks):Nt, (Nt-Nchecks):Nt]) = 99999
dim(Kaug)
str(Kaug)

```

Convert the K matrix to a sparse form. This takes a while.
```{r}
Nelem = N + ((Nt*(Nt-1))/2) #number of elements is num diag + num off-diag on HALF the matrix
Ksparse = data.frame(r = rep(NA, Nelem), c= rep(NA, Nelem), val = rep(NA, Nelem))
Index = 1
for (i in 1:Nt){
  for (j in i:Nt){
    Ksparse[Index,] = c(j, i, Kaug[i,j])
    Index = Index + 1
  }
}   
Ksparse = Ksparse[order(Ksparse[,1], Ksparse[,2]),]
```

Write the files out for asreml to use
```{r}
write.table(ped2, file = "ReFusC3_C4_fake_ped.txt", quote = F, row.names = F, col.names = F)

name_vector = ped2[,"line", drop = F]
write.table(name_vector, file = "ReFUSC3_C4_K_line_order.txt", quote = F, row.names = F, col.names = F)

write.table(Ksparse, file = "ReFusC3_C4_K.grm", quote = F, row.names=F, col.names = F)
```

Quick check on the K matrix diagonals
```{r}
Kmat = as.matrix(K)
summary(diag(Kmat))
hist(diag(Kmat))
```

Check the off-diagonals
```{r}
Npairs = (N*(N-1))/2
Koff = vector(length = Npairs)
index = 1

for (i in (1:(N-1))){
  for (j in (i+1):N){
  Koff[index] = K[i,j]
  index = index + 1
}}

summary(Koff)
hist(Koff)
```

